### 2017-05-15 Kokkos_PG_Compilation.tex
SOme more updates

In order to compile Kokkos a C++11 compliant compiler is needed. 
For an up to date list of compilers which are tested on a nightly basis, please refer to the README on the github repository. 
At the time of writing supported compilers include: 
\begin{lstlisting}
  GCC   4.7.2, 4.8.x, 4.9.x, 5.1.x, 5.3.x
  Intel 14.x, 15.x, 16.x, 17.x 
  Clang 3.5.1, 3.6.x, 3.7.x, 3.8.x, 3.9.x; 
  Cuda  6.5, 7.0, 7.5, 8.0; 
  XL 13.3;
\end{lstlisting}

@@

\hline    
    KOKKOS\_ARCH (IN): & What backend Architecture to build for. \\
      \hspace{0.5cm}Options: & KNL,KNC,SNB,HSW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
               Maxwell50,ARMv8,BGQ,Power7,Power8 \\
      \hspace{0.5cm}Default: & None -- this means no particular architecture flags
               are set. \\

\hline
   KOKKOS\_USE\_TPLS (IN): & Enable optional third party libraries. \\
      \hspace{0.5cm}Options: & hwloc,librt \\
      \hspace{0.5cm}Default: & \\

@@

Any Kokkos application compiled for CUDA embeds CUDA code via template meta-programming.
Thus, the whole application must be built with a CUDA-capable compiler.
(At the moment, the only such compiler is NVIDIA's NVCC.)
More precisely, every compilation unit containing a Kokkos kernel or a function called from a Kokkos kernel has to be compiled with a CUDA-capable compiler. 
This includes files containing \lstinline|Kokkos::View| allocations, which call an initialization kernel.

>>>>

In order to compile Kokkos a C++11 compliant compiler is needed. 
For an up to date list of compilers which are tested on a nightly basis, please refer to the README on the github repository. 
At the time of writing supported compilers include: 
\begin{lstlisting}
  GCC   4.7.2, 4.8.x, 4.9.x, 5.1.x, 5.3.x
  Intel 14.x, 15.x, 16.x, 17.x 
  Clang 3.5.1, 3.6.x, 3.7.x, 3.8.x, 3.9.x; 
  Cuda  7.0, 7.5, 8.0; 
  XL 13.3;
\end{lstlisting}

@@

\hline    
    KOKKOS\_ARCH (IN): & What backend Architecture to build for. \\
      \hspace{0.5cm}Options: & KNL,KNC,SNB,HSW,BDW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
               Maxwell50,Pascal60,Pascal61,ARMv8,ARMv81,ARMv8-ThunderX,BGQ,Power7,Power8 \\
      \hspace{0.5cm}Default: & None -- this means no particular architecture flags
               are set. \\

\hline
   KOKKOS\_USE\_TPLS (IN): & Enable optional third party libraries. \\
      \hspace{0.5cm}Options: & hwloc,librt,experimental\_memkind \\
      \hspace{0.5cm}Default: & \\

@@

Any Kokkos application compiled for CUDA embeds CUDA code via template meta-programming.
Thus, the whole application must be built with a CUDA-capable compiler.
(At the moment, the only such compilers are NVIDIA's NVCC and Clang 4.0 [not released yet at time of writing].)
More precisely, every compilation unit containing a Kokkos kernel or a function called from a Kokkos kernel has to be compiled with a CUDA-capable compiler. 
This includes files containing \lstinline|Kokkos::View| allocations, which call an initialization kernel.

### 2017-05-15 Kokkos_PG_HierarchicalParallelism.tex
SOme more updates

You should use Hierarchical Parallelism in particular in a number of cases:
\begin{enumerate}
\item Non-tightly nested loops: Hierarchical Parallelism allows you to expose more parallelism.
\item Data gather + reuse: If you gather data for a particular iteration of an outer loop, and then repeatiptly use it in an inner loop, Hierarchical Parallelism with scratch memory may match the usecase well.
\item Force Cache Blocking: Using Hierarchical Parallelism forces a developer into algorithmic choices which are good for cache blocking. This can sometimes lead to better performing algorithms, than simple flat parallelism.
\end{enumarate}

@@

\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}:

@@

\begin{lstlisting}
tyepdef Kokkos::DefaultExecutionSpace::scratch_memory_space
  ScratchSpace;
// Define a view type in ScratchSpace
typedef Kokkos::View<int*[4],ScratchSpace,Kokkos::MemoryTraits<Kokkos::Unmanaged>> shared_int_2d;

// Get the size of the shared memory allocation
size_t shared_size = shared_int_2d::shmem_size(team_size);
Kokkos::parallel_for(Kokkos::TeamPolicy<>(league_size,team_size),
                     KOKKOS_LAMBDA ( member_type team_member) {
  // Get a view allocated in team shared memory.
  // The constructor takes the shared memory handle and the 
  // runtime dimensions
  shared_int_2d A(team_member.team_shmem(), team_member.team_size());
  ...

});
\end{lstlisting}

@@

Kokkos provides the \lstinline|Kokkkos::single(Policy,Lambda)| function for this case.
It currently accepts two policies:
\begin{itemize}
\item \lstinline|Kokkos::PerTeam| restricts execution of the lambda's
  body to once per team
\item \lstinline|Kokkos::PerThread| restricts execution of the
  lambda's body to once per thread (that is, to only one vector lane
  in a thread)
\end{itemize}
The \lstinline|single| function takes a lambda as its second argument.
That lambda takes zero arguments, 
so its body must perform side effects in order to have an effect.
It must always be correct for the lambda to capture variables by value
(\lstinline|[=]|, not \lstinline|[&]|).
Thus, if the lambda captures by reference,
it must \emph{not} modify variables that it has captured by reference.

@@

>>>>

You should use Hierarchical Parallelism in particular in a number of cases:
\begin{enumerate}
\item Non-tightly nested loops: Hierarchical Parallelism allows you to expose more parallelism.
\item Data gather + reuse: If you gather data for a particular iteration of an outer loop, and then repeatiptly use it in an inner loop, Hierarchical Parallelism with scratch memory may match the usecase well.
\item Force Cache Blocking: Using Hierarchical Parallelism forces a developer into algorithmic choices which are good for cache blocking. This can sometimes lead to better performing algorithms, than simple flat parallelism.
\end{enumerate}

@@

\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}

@@

\begin{lstlisting}
tyepdef Kokkos::DefaultExecutionSpace::scratch_memory_space
  ScratchSpace;
// Define a view type in ScratchSpace
typedef Kokkos::View<int*[4],ScratchSpace,
          Kokkos::MemoryTraits<Kokkos::Unmanaged>> shared_int_2d;

// Get the size of the shared memory allocation
size_t shared_size = shared_int_2d::shmem_size(team_size);
Kokkos::parallel_for(Kokkos::TeamPolicy<>(league_size,team_size),
                     KOKKOS_LAMBDA ( member_type team_member) {
  // Get a view allocated in team shared memory.
  // The constructor takes the shared memory handle and the 
  // runtime dimensions
  shared_int_2d A(team_member.team_shmem(), team_member.team_size());
  ...  
});
\end{lstlisting}

@@

Kokkos provides the \lstinline|Kokkkos::single(Policy,Lambda)| function for this case.
It currently accepts two policies:
\begin{itemize}
\item \lstinline|Kokkos::PerTeam| restricts execution of the lambda's
  body to once per team
\item \lstinline|Kokkos::PerThread| restricts execution of the
  lambda's body to once per thread (that is, to only one vector lane
  in a thread)
\end{itemize}
The \lstinline|single| function takes a lambda as its second argument.
That lambda takes zero arguments or one argument by reference. 
If it takes no argument, its body must perform side effects in order to have an effect.
If it takes one argument, the final value of that argument is broadcast to every executor on the level: 
i.e. every vectorlane of the thread, or every thread (and vector lane) of the team.  
It must always be correct for the lambda to capture variables by value
(\lstinline|[=]|, not \lstinline|[&]|).
Thus, if the lambda captures by reference,
it must \emph{not} modify variables that it has captured by reference.

@@

Here is an example of using the broadcast capabilities to determine the start offset for a team
in a buffer:

Kokkos::View<int> offset("Offset");
offset() = 0;

  parallel_reduce (TeamThreadRange (thread, 100),
    KOKKOS_LAMBDA (const int& i, int& lsum) {
      if(...) lsum++;
  });
  Kokkos::single (PerTeam (thread), [=] (int& my_offset) {
    my_offset = Kokkos::atomic_fetch_add(&offset(),lsum);
  });
  ...
\end{lstlisting}

To further illustrate the "parallel region" semantics of the team execution consider the following code:

\begin{lstlisting}
using Kokkos::parallel_reduce;
using Kokkos::TeamThreadRange;
using Kokkos::TeamPolicy;

parallel_reduce(TeamPolicy<>(N,team_size),
  KOKKOS_LAMBDA (const member_type& teamMember, int& lsum) {
    int s = 0;
    for(int i = 0; i<10; i++) s++;
    lsum += s;
},sum);
\end{lstlisting}

In this example \lstinline{sum} will contain the value \lstinline{N * team_size * 10}.
Every thread in each team will compute \lstinline{s=10} and then contribute it to the sum.

Lets go one step further and add a nested \lstinline{parallel_reduce}. 
By choosing the loopbound to be \lstinline{team_size} every thread still only runs
once through the inner loop.

\begin{lstlisting}
using Kokkos::parallel_reduce;
using Kokkos::TeamThreadRange;
using Kokkos::TeamPolicy;

parallel_reduce(TeamPolicy<>(N,team_size),
  KOKKOS_LAMBDA (const member_type& teamMember, int& lsum) {

  int s = 0;
  parallel_reduce(TeamThreadRange(teamMember, team_size),
    [=] (const int k, int & inner_lsum) {
    int inner_s = 0;
    for(int i = 0; i<10; i++) inner_s++;
    inner_lsum += inner_s;
  },s);
  lsum += s;
},sum);
\end{lstlisting}

The answer in this case is neverless \lstinline{N * team_size * team_size * 10}.
Each thread computes \lstinline{inner_s = 10}.
But all threads in the team combine their results to compute a \lstinline{s} value of \lstinline{team_size * 10}.
Since every thread in each team contributes that value to the global sum, we arrive at the final value of \lstinline{N * team_size * team_size * 10}. 
If the intended goal was for each team to only contribute \lstinline{s} once to the global sum, 
the contribution should have been protected with a \lstinline{single} clause.

### 2017-05-15 Kokkos_PG_MachineModel.tex
SOme more updates

Kokkos is comprised of two orthogonal aspects. The first of these is 
an underlying {\em abstract machine model} which describes fundamental
concepts required for the development of future portable and performant
high performance computing applications; the second is a concrete instantiation
of the programming this model written in C++, which allows programmers
to write to the concept machine model. It is important to treat these
two aspects to Kokkos as distinct entities because the underlying model
being used by Kokkos could, in the future, be instantiated in additional
languages beyond C++ yet the algorithmic specification would remain
valid.

@@

Kokkos assumes an {\em abstract machine model} for the design of
future shared-memory computing architectures. The model (shown in
Figure~\ref{fig:kokkosnode}) assumes that there may be multiple
execution units in a compute node. For a more general discussion
of abstract machine models for Exascale computing the reader should
consult reference~\cite{cal_amm}. In the figure shown here,
we have elected to
show two different types of compute units -- one which represents
multiple latency-optimized cores, similar to contemporary processor
cores, and a second source of compute in the form of an off die
accelerator. Of note is that the processor and accelerator each 
have distinct memories, each with unique performance properties,
that are accessible across the node ({\em i.e.} the memory is
reachable or {\em shared} by all execution units. The specific
layout shown in Figure~\ref{fig:kokkosnode} is an instantiation
of the Kokkos abstract machine model used to describe the potential
for multiple types of compute engines and memories within a single
node. In future systems there may be a range of execution engines
which are used in the node ranging from a single type of core
as in many/multi-core processors found today through to an range
of execution units where many-core processors may be joined to
numerous types of accelerator cores. In order to ensure portability
to the potential range of nodes an abstraction of the
compute engines and available memories are required.

@@

An {\em instance} of an execution space is a specific instantiation
of an execution space to which a programmer can target parallel
work. By means of example, an execution space might be used
to describe a multi-core processor. In this example, the
execution space contains several homogeneous cores which share
some logical grouping. In a program written to the Kokkos model,
an instance of this execution space would be made available
on which parallel kernels could be executed. As a second example,
if we were to add a GPU to the multi-core processor a second
execution space type is available in the system, the 
application programmer would then have two 
execution space instances available to select from. The
important consideration here is that the method of
compiling code for different execution spaces and the
dispatch of kernels to instances is abstracted by the Kokkos
framework allowed application programmers to be free from 
written algorithms in hardware specific languages or
requiring the programmer to understand the restrictions
required for parallel execution in the execution space.

@@

The multiple types of memory which will become available in future
computing nodes are abstracted by Kokkos through {\em memory
spaces}. Each memory space provides a finite storage capacity
at which data structures can be allocated and accessed. In
order to provide portable allocations and performant access,
Kokkos requires data to be allocated as a $k$-dimensioned array,
where $k$ must be greater than equal to zero (a zero array being
a scalar). Each dimension of the array can be either fixed in
size at compile time or dynamically sized with size resolution
performed during execution.

@@

Although atomic operations are logically an operation
on memory, they are performed by compute units and so
are a property of execution spaces in the Kokkos framework.
Execution spaces are not guaranteed to provide atomic
operations support but attempts to use them on spaces
which do not support atomic access will result in 
a compile time error.

@@

Memory consistency models are a complex topic in and of themselves
and usually rely on complex operations associated with hardware
caches or memory access coherency (for more information 
see reference~\cite{handp_hardware}. 
Kokkos, does not {\em require} caches to be present
in hardware and so assumes an extremely weak memory consistency
model. In the Kokkos model, the programmer should not assume any
specific ordering of memory operations being isused by a
kernel. This has the potential to create race conditions between
memory operations if these are not appropriately protected. In
order to provide a guaranteed that memory are completed, Kokkos
provides a {\em fence} operation which forces the compute 
engine to complete all outstanding memory operations before
any new ones can be issued. With appropriate use of fences,
programmers are thereby able to ensure that guarantees can
be made as to when data will {\em definitely} have been 
written to memory.

>>>>

Kokkos is comprised of two orthogonal aspects. The first of these is 
an underlying {\em abstract machine model} which describes fundamental
concepts required for the development of future portable and performant
high performance computing applications; the second is a concrete instantiation
of the programming model written in C++, which allows programmers
to write to the concept machine model. It is important to treat these
two aspects to Kokkos as distinct entities because the underlying model
being used by Kokkos could, in the future, be instantiated in additional
languages beyond C++ yet the algorithmic specification would remain
valid.

@@

Kokkos assumes an {\em abstract machine model} for the design of
future shared-memory computing architectures. The model (shown in
Figure~\ref{fig:kokkosnode}) assumes that there may be multiple
execution units in a compute node. For a more general discussion
of abstract machine models for Exascale computing the reader should
consult reference~\cite{cal_amm}. In the figure shown here,
we have elected to
show two different types of compute units -- one which represents
multiple latency-optimized cores, similar to contemporary processor
cores, and a second source of compute in the form of an off die
accelerator. Of note is that the processor and accelerator each 
have distinct memories, each with unique performance properties,
that may or may not be accessible across the node ({\em i.e.} the memory may be
reachable or {\em shared} by all execution units, but specific 
memory spaces may also be only accessible by specifc execution units). 
The specific layout shown in Figure~\ref{fig:kokkosnode} is an instantiation
of the Kokkos abstract machine model used to describe the potential
for multiple types of compute engines and memories within a single
node. In future systems there may be a range of execution engines
which are used in the node ranging from a single type of core
as in many/multi-core processors found today through to an range
of execution units where many-core processors may be joined to
numerous types of accelerator cores. In order to ensure portability
to the potential range of nodes an abstraction of the
compute engines and available memories are required.

@@

An {\em instance} of an execution space is a specific instantiation
of an execution space to which a programmer can target parallel
work. By means of example, an execution space might be used
to describe a multi-core processor. In this example, the
execution space contains several homogeneous cores which share
some logical grouping. In a program written to the Kokkos model,
an instance of this execution space would be made available
on which parallel kernels could be executed. As a second example,
if we were to add a GPU to the multi-core processor a second
execution space type is available in the system, the 
application programmer would then have two 
execution space instances available to select from. The
important consideration here is that the method of
compiling code for different execution spaces and the
dispatch of kernels to instances is abstracted by the Kokkos
model. 
This allows application programmers to be free from 
writing algorithms in hardware specific languages.

@@

The multiple types of memory which will become available in future
computing nodes are abstracted by Kokkos through {\em memory
spaces}. Each memory space provides a finite storage capacity
at which data structures can be allocated and accessed. 
Differnt memory space types have different characteristics with respect
to accesibility from execution spaces as well as their performance
characteristics.

@@

@@

Memory consistency models are a complex topic in and of themselves
and usually rely on complex operations associated with hardware
caches or memory access coherency (for more information 
see reference~\cite{handp_hardware}. 
Kokkos, does not {\em require} caches to be present
in hardware and so assumes an extremely weak memory consistency
model. In the Kokkos model, the programmer should not assume any
specific ordering of memory operations being isued by a
kernel. This has the potential to create race conditions between
memory operations if these are not appropriately protected. In
order to provide a guaranteed that memory are completed, Kokkos
provides a {\em fence} operation which forces the compute 
engine to complete all outstanding memory operations before
any new ones can be issued. With appropriate use of fences,
programmers are thereby able to ensure that guarantees can
be made as to when data will {\em definitely} have been 
written to memory.

### 2017-05-15 Kokkos_PG_ProgrammingModel.tex
SOme more updates

@@

Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
and share a ``scratch pad'' memory which they may use for temporary storage.
Note that the not all forms of synchronization mechanism are legal in Kokkos, in particular
implementing ``spin-locks'' for threads in a team may result in dead locks. 
In Kokkos there is no forward progress guarantee for threads, that means that a single thread spinning
on a lock acquired by another thread of the team may use 100\% of the compute engines resources.
Calling the explicit Kokkos team barrier is the only safe way to synchronize threads in a team.

@@

Memory Spaces are the places {\it Where} data resides.
They specify physical location of data as well as certain access characteristics. 
Different physical locations correspond to things such as high bandwidth memory, on die scratch memory or non-volatile bulk storage.
Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from Host and GPU. 
Memory Spaces also can be used to express remote memory locations.
Furthermore they encapsulate functionality such as consistency control and persistence scopes.

@@

Layouts express the mapping from logical (or algorithmical) indicies to address offset for a data allocation. 
By adopting appropriate layouts for memort structures an application can optimise data access patterns in a given algorithm.
If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts) an architecture dependent optimisation can be performed.

>>>>

\begin{figure}[h]
\begin{center}
\includegraphics[clip, trim=1cm 5.5cm 12cm 2.5cm, width=\textwidth]{figures/kokkos-abstractions.pdf}
\caption{The core abstractions of the Kokkos Programming Model}
\label{fig:kokkosabstractions}
\end{center}
\end{figure}

@@

Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
and share a ``scratch pad'' memory which they may use for temporary storage.
Note that not all forms of synchronization mechanism are legal in Kokkos, in particular
implementing ``spin-locks'' for threads in a team may result in dead locks. 
In Kokkos there is no forward progress guarantee for threads, that means that a single thread spinning
on a lock acquired by another thread of the team may use 100\% of the compute engines resources.
Calling the explicit Kokkos team barrier is the only safe way to synchronize threads in a team.

@@

Memory Spaces are the places {\it Where} data resides.
They specify physical location of data as well as certain access characteristics. 
Different physical locations correspond to things such as high bandwidth memory, on die scratch memory or non-volatile bulk storage.
Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from Host and GPU. 
Memory Spaces also could in the futrue also be used to express remote memory locations.
Furthermore they encapsulate functionality such as consistency control and persistence scopes.

@@

Layouts express the mapping from logical (or algorithmical) indicies to address offset for a data allocation. 
By adopting appropriate layouts for memory structures an application can optimise data access patterns in a given algorithm.
If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts) an architecture dependent optimisation can be performed.

### 2017-05-15 Kokkos_PG_Subviews.tex
SOme more updates

To take a subview of a View, use the \texttt{Kokkos::subview}
function.  This function is overloaded for all different kinds of
Views and index ranges.  For example, the following code is equivalent
to the above example \texttt{A(2:4, 3:7)}:
%%
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**> A ("A", N0, N1);

@@

A subview has the same reference count as its parent View, so the
parent View won't be deallocated before all subviews go away.  Every
subview is also a View.  This means that you may take a subview of a
subview.

@@

Note the use of the C++11 keyword \texttt{auto} in the above example.
A subview may have a different type than its parent View.  For
instance, if \lstinline!A! has \lstinline!LayoutRight!,
\lstinline!A_sub!  has \lstinline!LayoutStride!.  A subview of an
entire row of \lstinline!A! has \lstinline!LayoutRight! as well, but a
subview of an entire column of \lstinline!A! has
\lstinline!LayoutStride!.  If you assign the result of the
\texttt{subview} function to the wrong type, Kokkos will emit a
compile-time error.  The best way to avoid this is to use the C++11
\texttt{auto} keyword to let the compiler deduce the correct type for
you.

>>>>

To take a subview of a View, you can use the \texttt{Kokkos::subview}
function.  This function is overloaded for all different kinds of
Views and index ranges.  For example, the following code is equivalent
to the above example \texttt{A(2:4, 3:7)}:
%%
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**> A ("A", N0, N1);

@@

A subview has the same reference count as its parent \lstinline!View!, so the
parent \lstinline!View! won't be deallocated before all subviews go away.  Every
subview is also a \lstinline!View!.  This means that you may take a subview of a
subview.

Another way of getting a subview is throught the appropriate \lstinline{View} constructor.
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**> A ("A", N0, N1);

View<double**,LayoutStride> A_sub(A,make_pair(2,4),make_pair(3,7));
\end{lstlisting}
For this usage you must know the layouttype of the subview. 
On the positive side, such a direct construction is generally a bit cheaper than
through the \lstinline!Kokkos::subview! function.

@@

Note the use of the C++11 keyword \texttt{auto} in the above example.
A subview may have a different type than its parent View.  For
instance, if \lstinline!A! has \lstinline!LayoutRight!,
\lstinline!A_sub!  has \lstinline!LayoutStride!.  A subview of an
entire row of \lstinline!A! has \lstinline!LayoutRight! as well, but a
subview of an entire column of \lstinline!A! has
\lstinline!LayoutStride!.  If you assign the result of the
\texttt{subview} function to the wrong type, Kokkos will emit a
compile-time error.  The easiest way to avoid this is to use the C++11
\texttt{auto} keyword to let the compiler deduce the correct type for
you. That said \lstinline!auto! comes with its own cost. 
It generally is more expansive for compilers to deal with \lstinline!auto! than with 
explicitly known types.
